Models

### 2021-06-03 analysis/05_model_metrics.Rmd
Fix scaling so all models converge, all plots in one figure with a legend.

```{r}
# model for overall trend (random effect w/varying intercept)
# metric_mean_fit <- model_data %>%
#   filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
#   group_nest(metric) %>%
#   mutate(model = map(data, ~lmer(log(value + 1) ~ year_scaled + (1|study_id/rarefyID), data = .x, REML = FALSE,
#                   control = lmerControl(optCtrl = list(maxit = 1e9,
#                                                     maxfun = 1e9,
#                                                     xtol_abs = 1e-11,
#                                                     ftol_abs = 1e-11,
#                                                     maxeval = 1e9)))),
#          null_model = map(data, ~lmer(logvalue ~ (1|study_id/rarefyID), data = .x, REML = FALSE,
#                   control = lmerControl(optCtrl = list(maxit = 1e9,
#                                                     maxfun = 1e9,
#                                                     xtol_abs = 1e-11,
#                                                     ftol_abs = 1e-11,
#                                                     maxeval = 1e9)))),
#            
#            test = map2(model, null_model, anova),
#            test_res = map(test, tidy),
#            coef = map(model, tidy)
#            ) %>% 
#   unnest(cols = c(test_res)) %>%
#   select(metric, model, coef, p.value) %>%
#   filter(!is.na(p.value)) %>%
#   unnest(cols = c(coef)) %>%
#   filter(term == "year_scaled") %>%
#   select(-c(effect, group, term, statistic))

SES_rand_slope <- model_data %>%
  #filter(protected_area == "FALSE") %>%
  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis", )) %>%
  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "Jaccard_next")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
           #                       data = .x)),
           # test = map2(model, null_model, anova),
           # test_res = map(test, tidy),
           coef = map(model, tidy),
           ) %>% 
  select(metric, model, coef) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

@@

metric_rand_slope <- model_data %>%
  #filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
  filter(metric %in%  c("S")) %>%
  #it wants its own scaling to converge for some reason
  mutate(year_scaled = scale(year)) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x, REML = TRUE)),
           coef = map(model, tidy)
           ) %>% 
  select(metric, model, coef) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

#check that they all used the same number of observations
map(metric_rand_slope$model, nobs)

# model_comp <- metric_mean_fit %>%
#   select(metric, no_slope_estimate = estimate, no_slope_pval = p.value) %>%
#   left_join(metric_mean_fit_slope %>% 
#   select(metric, slope_estimate = estimate, slope_pval = p.value ))

metric_model_table <- bind_rows(select(SES_rand_slope, -model), select(metric_rand_slope, -model))
```

@@

taxa_slope_model_S <- model_data %>% 
  filter(metric == "S") %>%
  lmer(logvalue ~ year_scaled + taxa + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
                                 data = ., REML = TRUE)
```
Realm:
```{r}
realm_slope_model <- model_data %>%
  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + realm + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
                                 data = .x, REML = TRUE)),
           coef = map(model, tidy)) %>%
  select(metric, coef) %>%
  unnest(cols = c(coef)) %>%
  #filter(term == "year_scaled") %>%
  filter(effect == "fixed") %>%
  select(-c(effect, group, statistic))

realm_slope_model_S <- model_data %>% 
  filter(metric == "S") %>%
  lmer(logvalue ~ year_scaled + realm + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
                                 data = ., REML = TRUE)
```
Climate:
```{r}
climate_slope_model <- model_data %>%
  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + climate + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
                                 data = .x, REML = TRUE)),
           coef = map(model, tidy)) %>%
  select(metric, coef) %>%
  unnest(cols = c(coef)) %>%
  #filter(term == "year_scaled") %>%
  filter(effect == "fixed") %>%
  select(-c(effect, group, statistic))

climate_slope_model_S <- model_data %>% 
  filter(metric == "S") %>%
  lmer(logvalue ~ year_scaled + climate + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
                                 data = ., REML = TRUE)
```

@@

study_slopes <- pmap_dfr(bind_rows(select(SES_rand_slope, metric, model), select(metric_rand_slope,  metric, model)), get_study_slopes)

@@

Summaries of study slopes
```{r}
#First - how many studies had a significant trend for each metric? 
study_slopes %>% 
  filter(sig == TRUE) %>% 
  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
  group_by(metric) %>% 
  summarise(n_study_id = n_distinct(study_id)) %>%
  mutate(percent = n_study_id/54)

@@

#### Analizing slopes
Pairwise correlation analysis
```{r}
slope_corr <- corr.test(slopes_wide %>% ungroup() %>% select(starts_with("SES"), "S", "Jaccard_base"))

@@

```{r}
plot_metric <- function(metric_name, ylabel, yval, data, model_coef){

@@

  trend_plot <- data %>%
    left_join(pred, by = "year_scaled") %>%
    #left_join(metadata %>% select(study_id, climate)) %>%
    ggplot(aes(x = year, y = !!sym(yval))) +
    # geom_smooth(aes(color = climate, group = rarefyID), method = "glm", se = FALSE) +
    # geom_point(aes(color = climate, group = rarefyID)) +
    geom_point(aes(group = rarefyID) ,color = "grey", size = 0.50) +
    geom_smooth(aes(color = climate, group = study_id), method = "glm", se = FALSE, size = 0.7) +
    #geom_smooth(color = "black", method = "glm", se = FALSE) +
    geom_line(mapping = aes(y = pred)) +
    theme(legend.position = "none") +
    theme_classic() +
    ylab(ylabel) +
    labs(color = "Climate") +
    scale_color_manual(values = c("#32a251", 
                "#ff7f0f", "#39737c", "#b85a0d", "#ffd94a"))

@@

model_coef <- bind_rows(SES_rand_slope, metric_rand_slope) %>%
  select(metric, model) %>%
   mutate(coef = map(model, tidy)) %>% 
  select(-model) %>% 
  unnest(cols = c(coef))

plot_map <- tibble(metric_name = c("S", "SES_FRic", "Jaccard_base"), 
           ylabel = c("log(S)", "Functional Richness", "Jaccard"),
           yval = c("logvalue", "value", "value"))
m_plots <- pmap(plot_map, plot_metric, data = model_data, model_coef = model_coef)

plot_join <- plot_grid(plotlist=m_plots, labels = c("A", "B", "C"), nrow = 3)
ggsave2("3met_long.jpeg")
```

>>>>

```{r}
SES_rand_slope <- model_data %>%
  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
  group_nest(metric) %>%
  mutate(#year_center = map(data, ~scale(.x$year)),
         model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
           #                       data = .x)),
           # test = map2(model, null_model, anova),
           # test_res = map(test, tidy),
           coef = map(model, tidy),
           ) %>% 
  select(metric, model, coef) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

@@

#these two metrics are cranky and want their own year scaling to converge

#Jaccard Base
jacc_base_mod <-  model_data %>%
  filter(metric == ("Jaccard_base")) %>%
  mutate(year_scaled = scale(year)) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           coef = map(model, tidy),
           ) %>% 
  select(metric, model, coef) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

# S model
s_mod <- model_data %>%
  filter(metric %in%  c("S")) %>%
  #it wants its own scaling to converge for some reason
  mutate(year_scaled = scale(year)) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x, REML = TRUE)),
           coef = map(model, tidy)
           ) %>% 
  select(metric, model, coef) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

metric_model_table <- bind_rows(SES_rand_slope, s_mod, jacc_base_mod)
```

@@

taxa_slope_model_S <- model_data %>% 
  mutate(year_scaled = scale(year)) %>% 
  filter(metric == "S") %>%
  lmer(logvalue ~ year_scaled + taxa + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
                                 data = ., REML = TRUE)
```
Realm:
```{r}
realm_slope_model <- model_data %>%
  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + realm + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
                                 data = .x, REML = TRUE)),
           coef = map(model, tidy)) %>%
  select(metric, coef) %>%
  unnest(cols = c(coef)) %>%
  #filter(term == "year_scaled") %>%
  filter(effect == "fixed") %>%
  select(-c(effect, group, statistic))

realm_slope_model_S <- model_data %>% 
  mutate(year_scaled = scale(year)) %>% 
  filter(metric == "S") %>%
  lmer(logvalue ~ year_scaled + realm + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
                                 data = ., REML = TRUE)
```
Climate:
```{r}
climate_slope_model <- model_data %>%
  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + climate + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
                                 data = .x, REML = TRUE)),
           coef = map(model, tidy)) %>%
  select(metric, coef) %>%
  unnest(cols = c(coef)) %>%
  #filter(term == "year_scaled") %>%
  filter(effect == "fixed") %>%
  select(-c(effect, group, statistic))

climate_slope_model_S <- model_data %>% 
  mutate(year_scaled = scale(year)) %>% 
  filter(metric == "S") %>%
  lmer(logvalue ~ year_scaled + climate + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
                                 data = ., REML = TRUE)
```

@@

study_slopes <- pmap_dfr(select(metric_model_table,  metric, model), get_study_slopes)

@@

Summaries of study slopes
```{r}
#First - how many studies had a significant trend for each metric? 
study_slopes %>% 
  filter(sig == TRUE) %>% 
  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
  group_by(metric) %>% 
  summarise(n_study_id = n_distinct(study_id)) %>%
  mutate(percent = n_study_id/53)

@@

#### Analizing slopes
Pairwise correlation analysis
```{r}
library(psych)
#slope_corr <- corr.test(slopes_wide %>% ungroup() %>% select(starts_with("SES"), "S", "Jaccard_base"))

@@

```{r}
library(cowplot)
library(RColorBrewer)

plot_metric <- function(metric_name, ylabel, yval, xlabel, data, model_coef, color_list){

  if(is.na(xlabel)) xlabel <- NULL

@@

  trend_plot <- data %>%
    left_join(pred, by = "year_scaled") %>%
    #left_join(metadata %>% select(study_id, climate)) %>%
    ggplot(aes(x = year, y = !!sym(yval))) +
    # geom_smooth(aes(color = climate, group = rarefyID), method = "glm", se = FALSE) +
    # geom_point(aes(color = climate, group = rarefyID)) +
    geom_point(aes(group = rarefyID) ,color = "grey", size = 0.50) +
    geom_smooth(aes(color = climate, group = study_id), method = "glm", se = FALSE, size = 0.7) +
    #geom_smooth(color = "black", method = "glm", se = FALSE) +
    geom_line(mapping = aes(y = pred)) +
    theme_classic() +
    theme(axis.title = element_text(size=7), legend.position = "none") +
    ylab(ylabel) +
    xlab(xlabel) +
    labs(color = "Climate") +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 6)) +
    scale_colour_manual(values = color_list, drop = FALSE)

    #scale_color_manual(values = c("#32a251", 
               # "#ff7f0f", "#39737c", "#b85a0d", "#ffd94a"))

@@

model_coef <- metric_model_table %>%
  select(metric, model) %>%
   mutate(coef = map(model, tidy)) %>% 
  select(-model) %>% 
  unnest(cols = c(coef))

pal <- brewer.pal(n = 5, name = "Dark2")
group_colors <- c("Global" = pal[1], "Polar/Temperate" = pal[2], "Temperate/Tropical" = pal[3], "Tropical" = pal[4], "Temperate" = pal[5], "Overall Mean" = "#000000")

plot_map <- tibble(metric_name = c("S", "Jaccard_base", "SES_FRic", "SES_FDiv", "SES_FEve"), 
           ylabel = c("log(S)", "Jaccard", "Functional Richness SES", "Functional Divergence SES", "Functional Evenness SES"),
           #ylabel = c("log(S)", "Jaccard", "FRic SES", "FDiv SES", "FEve SES"),
           yval = c("logvalue", "value", "value", "value", "value"),
           xlabel = c(NA, NA, NA, "year", "year"))
m_plots <- pmap(plot_map, plot_metric, data = model_data, model_coef = model_coef, color_list = group_colors)

plot_join <- plot_grid(plotlist=m_plots, labels = "AUTO", nrow = 3, align = 'vh',
                       hjust = -1, axis = "l", scale = 0.9)

leg_plot <- model_data %>%
  bind_rows(data.frame(climate = c("Overall Mean", "Overall Mean"), year = c(2019, 2020), value = c(1, 1))) %>%
  ggplot(aes(x = year, y = value)) +
  geom_point(color = "grey", size = 0.50) +
  geom_smooth(aes(color = climate, group = study_id), method = "glm", se = FALSE, size = 0.7) +
  theme_classic() +
  theme(axis.title = element_text(size=7)) +
  scale_colour_manual(values = group_colors, drop = FALSE)

######
group.colors <- c(A = "#333BFF", B = "#CC6600", C ="#9633FF", D = "#E2FF33", E = "#E3DB71")

simplePlot <- function(DT, tit) 
  ggplot(DT ,aes(x=Name, y=Value, fill=Group)) +
  geom_bar(stat="identity") + xlab("") + ggtitle(tit) +
  #Specify colours
  scale_fill_manual(values=group.colors)
#####

# extract the legend from one of the plots
legend <- get_legend(leg_plot)

# add the legend to the row we made earlier. Give it one-third of 
# the width of one plot (via rel_widths).
plot_join + draw_grob(legend,  2/3.3, -0.3, .3/3.3, 1, scale = 0.9)

ggsave2(here::here("figures", "3met_long.jpeg"))
```
